---
title: 'Lockdown Fitness: An FitBit Insight'
author: "Fredrick Boshe"
date: "20/04/2021"
output: 
  github_document: default
  rmarkdown::github_document: default
---

# An analysis of how the lockdown in Germany may have affected Fitness levels.

Being in Germany the past 15 months means a state of endless lockdowns to curb the spread of the virus. While these lockdowns have had some positive results in keeping the waves of spread low in Germany, they have undoubtedly taken physical and mental toll on people. The sense of time, purpose and social behaviors have been negatively impacted. While it would be hard to actively track of how one's body reacts to everything, technology has given us tools that can do this for us. 

I am the proud owner of a FitBit Charge 2, which is roughly 5 years old(!?) but still works like a charm. I graduated last year in September, left my student working position, changed cities and experienced my 3rd winter ever. All while going through a pandemic lockdown. Personally, i felt it took its toll, but i wanted to get better insights from my FitBit to analyze just how much the pattern was disrupted. 

Questions looking to be answered are: 

1. How did my daily fitness levels trend between October 2020 and March 2021?
2. How did my sleep quantity and quality change during the same time period? Is there any relationship between my fitness and my quality of sleep?
3. How was my heart-rate/health throughout this period? Are the perceptions related to/influenced by the demographics and success metrics of the school?

I utilize my FitBit data that i [exported](https://help.fitbit.com/articles/en_US/Help_article/1133.htm) to try and answer these questions and provide context to key observations. The data comes in JSON format, which can be tricky for some people to parse through. You can uses this [Github page](https://iccir919.github.io/fitbit-json-to-csv/) that helps convert your JSON files to CSV.


```{r setup, include=TRUE, results='hide', message=FALSE, warning=FALSE}

#Loading packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(ggcorrplot)
library(scales)
library(tufte)
library(lubridate)
library(streamgraph)
library(jsonlite)
library(emo)
library(zoo)


```

### Data: Importing and Cleaning 

```{r data, include=TRUE, results='hide'}
#Load datasets
## Method1: as CSV
alt<-read_csv("Dataset/Altitude.csv")
dis<-read_csv("Dataset/Distance.csv")
hrt1<-read_csv("Dataset/Heartrate1.csv")
hrt2<-read_csv("Dataset/Heartrate2.csv")
lam<-read_csv("Dataset/Light active minutes.csv")
mam<-read_csv("Dataset/Moderate active minutes.csv")
sam<-read_csv("Dataset/Sedentary minutes.csv")
vam<-read_csv("Dataset/Very active minutes.csv")
slp_sc<-read_csv("Dataset/sleep_score.csv")

## Method2: as JSON and turn it into a dataframe
path <- "./Dataset/Sleep/"
files <- dir(path, pattern = "*.json")

slp <- files %>%
  map_df(~fromJSON(file.path(path, .), flatten = TRUE))

#Date: Separate date and time (time is not relevant for this analysis)
#Separate date variable
alt<-alt%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)  

#Standardize dates and rename variables
alt$date<-mdy(alt$date)
alt<-alt%>%
  rename(altitude=Altitude) 

#Repeat process for all datasets. 

```

The "Date" column is designated as a character in all dataframes. It needs to be converted into in a "Date" data type. Since time is not of importance for this analysis, i decided to separate it, discard it and then convert the dates into correct format. 
Remember to remove columns you feel are not important to your analysis or duplicated across dataframes e.g. sleep_log_entry_id. Filter out low confidence recordings for heartbeat (FitBit defines 0 1nd 1 as low confidence recordings, times when FitBit did not record any values).

```{r clean, echo=FALSE}

#Separate date variable
dis<-dis%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

hrt1<-hrt1%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

hrt2<-hrt2%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

lam<-lam%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

mam<-mam%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

sam<-sam%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

vam<-vam%>%
  separate(Date,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time)

slp_sc<-slp_sc%>%
  separate(timestamp,
           into=c("date", "time"),
           sep=" ")%>%
  select(-time, -sleep_log_entry_id, -deep_sleep_in_minutes)

#Standardize dates and rename variables

dis$date<-mdy(dis$date)
dis<-dis%>%
  rename(distance=Distance) 


hrt1$date<-mdy(hrt1$date)
hrt1<-hrt1%>%
  rename(heartrate=Heartrate) 


hrt2$date<-mdy(hrt2$date)
hrt2<-hrt2%>%
  rename(heartrate=Heartrate) 


lam$date<-mdy(lam$date)
lam<-lam%>%
  rename(light_act_mins=`Light active minutes`) 


mam$date<-mdy(mam$date)
mam<-mam%>%
  rename(moderate_act_mins=`Moderate active minutes`) 

sam$date<-mdy(sam$date)
sam<-sam%>%
  rename(sedentary_mins=`Sedentary minutes`)

slp_sc$date<-ymd(slp_sc$date)

vam$date<-mdy(vam$date)
vam<-vam%>%
  rename(very_act_mins=`Very active minutes`) 

#Remove redundant columns
slp$dateOfSleep<-ymd(slp$dateOfSleep)

slp<-slp%>%
  select(-1, -3:-6, -13:-16,-17, -19, -20, -22, -23, 
         -25, -26, -28, -29, -31, -33)%>%
  rename(date=dateOfSleep)%>%
  as_tibble()

#Dealing with missing values to allow aggregation
colSums(is.na(slp))

slp<-slp%>%
  select(-type, -efficiency) #Not important

slp<-aggregate(. ~date, data = slp, sum, na.rm=TRUE, na.action=na.pass)


```

```{r clean2, include=TRUE}
###Aggregate the data by month (example)
alt<-aggregate(alt["altitude"], by=alt["date"], sum)
dis<-aggregate(dis["distance"], by=dis["date"], sum)

#Separate bpm and confidence columns
hrt<-rbind(hrt1,hrt2)
hrt<-hrt%>%
  separate(heartrate,
           into=c("bpm", "confidence"),
           sep=",")
hrt$bpm<-parse_number(hrt$bpm)
hrt$confidence<-parse_number(hrt$confidence)

#keep only high confidence recording (2 or 3)
hrt<-hrt%>%
  filter(confidence==2| confidence==3)
table(hrt$confidence)

#create min, max and avg bpm columns for each date
hrt<-hrt%>%
  group_by(date)%>%
  mutate(min_bpm=min(bpm),
         max_bpm=max(bpm),
         avg_bpm=mean(bpm))
hrt$avg_bpm<-as.integer(hrt$avg_bpm)

#remove bpm and confidence interval columns. Keep one observation per day
hrt<-hrt%>%
  select(-2:-3)%>%
  unique.data.frame()

#Aggregate sleep score by date
slp_sc<-slp_sc%>%
  group_by(date)%>%
  summarize(across(everything(), mean))
sum(duplicated(slp_sc$date))

#Keep data for the last 9 months only
con_date<-function(df){
  df<-df%>%
    filter(date>=today()- months(9))
  return(df)
}

alt<-con_date(alt)
dis<-con_date(dis)
hrt<-con_date(hrt)
lam<-con_date(lam)
mam<-con_date(mam)
sam<-con_date(sam)
slp<-con_date(slp)
slp_sc<-con_date(slp_sc)
vam<-con_date(vam)

#Merge the files. left join to hrt dataset
merge<-hrt%>%
  left_join(alt, by="date")%>%
  left_join(dis, by="date")%>%
  left_join(lam, by="date")%>%
  left_join(mam, by="date")%>%
  left_join(sam, by="date")%>%
  left_join(vam, by="date")%>%
  left_join(slp, by="date")%>%
  left_join(slp_sc, by="date")

#Check for duplicated dates
sum(duplicated(merge$date))

##Convert distance units from centimeters to meters
merge$distance<-merge$distance/100

```

## Day Insights: Activity levels
### Distance Covered

```{r distance, include=TRUE, out.height='100%', out.width='100%', warning=FALSE}
#create month and day columns
merge_dis<-merge%>%
  mutate(day=day(date),
         month=month(date),
         year=year(date))

#Create a column with month and year
merge_dis<-merge_dis%>%
  mutate(period=as.yearmon(paste(year, month), "%Y %m"))
merge_dis$period<-as.Date(merge_dis$period)#To allow mapping easier


#merge_dis_longer<-merge_dis%>%
  #pivot_longer(cols = c(day, month),
               #names_to="period",
               #values_to="value")

plot1<-merge_dis%>%
  drop_na(distance)%>%
  filter(month!=9)%>%
  ggplot(aes(x=day, y=period, fill=distance))+
  geom_tile(colour="white",size=0.25, na.rm = TRUE)+
  theme_minimal()+
  scale_y_date(date_breaks = 'months', date_labels = '%b-%Y', expand = c(0,0))+
  scale_fill_distiller(palette = "RdPu", direction=+1)+
  labs(x = "Day", 
       y = "Month",
       title = "Heatmap: Distance Covered by Date")+
  theme(plot.title = element_text(hjust = 0.5, size = 14, face="bold",
                                  margin = margin(t = 20, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 15, 
                                                    b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 0, r = 0, 
                                                    b = 15, l = 0)))


plot1

```
